---
title: "lab4_transcriptomics Myles Taillefer"
output: html_document
date: "2025-10-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
lapply(c(
  "docopt", "DT", "pheatmap","GenomicFeatures", "DESeq2",
  "edgeR", "systemPipeR", "systemPipeRdata", "BiocStyle", "GO.db", "dplyr",
  "tidyr", "stringr", "Rqc", "QuasR", "DT", "ape", "clusterProfiler","biomaRt","EnhancedVolcano"
), require, character.only = TRUE)
```



#Question 1

```{r q1, warning = FALSE, message = FALSE }
# get the path to the fastq files  
fq_path <- systemPipeRdata::pathList()$fastqdir

# import meta data
meta_data <- read.table(
  "/Users/mtaillefer00/Library/R/arm64/4.4/library/systemPipeRdata/extdata/param/targetsPE.txt", 
  header = TRUE)

#replace string with correct path
meta_data$FileName1 <- gsub('./data/',fq_path,meta_data$FileName1)
meta_data$FileName2 <- gsub('./data/',fq_path,meta_data$FileName2)

#display data table
library('DT')
DT::datatable(meta_data)
```

#Question 2
```{r q2.1, warning = FALSE, message = FALSE}

qc_res <- rqc(path = fq_path, pattern = '.fastq.gz', openBrowser = FALSE)
rqcCycleQualityBoxPlot(qc_res[0:12])+theme_bw()
rqcCycleQualityBoxPlot(qc_res[13:24])+theme_bw()
rqcCycleQualityBoxPlot(qc_res[25:36])+theme_bw()
```

Reads 27, 29 and 31 are significantly longer (samples: M1A, A1A, V1A).
The shortest number of cycles is ~75

```{r q2.2}
rqcCycleBaseCallsLinePlot(qc_res[0:12])
rqcCycleBaseCallsLinePlot(qc_res[13:24])
rqcCycleBaseCallsLinePlot(qc_res[25:36])
```
sample 34 has many N calls.

#Question 3
```{r 2.5, warning = FALSE, message = FALSE}
#create the output file names
meta_data$processed1 <- sub("^/Users/mtaillefer00/Library/R/arm64/4.4/library/systemPipeRdata/extdata/fastq/", "/Users/mtaillefer00/Documents/BIOL_4315_Lab_4_MT/processed_reads/", meta_data$FileName1) 
meta_data$processed1 <- sub("(\\.fastq\\.gz)$", "_processed\\1", meta_data$processed1) 

```

```{r q3, eval = FALSE}
#trim and filter the reads
for(i in 1:nrow(meta_data)){
  QuasR::preprocessReads(filename = meta_data$FileName1[i],outputFilename = meta_data$processed1[i], truncateEndBases = 3, Lpattern = 'GCCCGGGTAA', nBases = 1)
} 
```

#Question 4
```{r q4, echo=FALSE, message=FALSE, warning=FALSE, results = 'hide'}

##### 5.3 Alignments #####

#5.3.1 Indexing the reference genome

#create the output directory
dir.create("./hisat2_index", recursive = TRUE)

at_genome <- "data/GCF_000001735.4_TAIR10.1_genomic.fna"

#Use system2 to run hisat2 from within R

tryCatch({system2(command = "hisat2-build", 
                  args = c("-p","8", at_genome,
                           "./hisat2_index/tair10_1_index"),
                  stdout = TRUE, stderr = TRUE)}, error = function(e) {
                    paste("hisat2-build", "indexing failed with error:", e$message)
                  })

#5.3.2 Mapping the reads to the indexed genome

#question 4

dir.create("./sam_files",recursive = TRUE)
dir.create("./bam_files",recursive = TRUE)

meta_data$processedUnzip1 <- sub("\\.gz$", "", meta_data$processed1)
meta_data$sam1 <- paste('/Users/mtaillefer00/Documents/BIOL_4315_Lab_4_MT/sam_files/',meta_data$SampleName, sep = '')
meta_data$sam1 <- paste0(meta_data$sam1, '.sam')
meta_data$hisatLog <- paste('/Users/mtaillefer00/Documents/BIOL_4315_Lab_4_MT/bam_files/',meta_data$SampleName, sep = '')
meta_data$hisatLog <- paste0(meta_data$hisatLog, '_hisat2.log')
meta_data$bam1 <- paste('/Users/mtaillefer00/Documents/BIOL_4315_Lab_4_MT/bam_files/',meta_data$SampleName, sep = '')
meta_data$bam1 <- paste0(meta_data$bam1, '.bam')
meta_data$bam1sorted <- paste('/Users/mtaillefer00/Documents/BIOL_4315_Lab_4_MT/bam_files/',meta_data$SampleName, sep = '')
meta_data$bam1sorted <- paste0(meta_data$bam1sorted, '.sorted.bam')

for( i in 1:nrow(meta_data)){
  hisat2 <- tryCatch({system2("hisat2", 
                    args = c('-x', 'hisat2_index/tair10_1_index', 
                             '-U',meta_data$processedUnzip1[i],
                             '-p', '8',
                             '-S', meta_data$sam1[i]),
                    stdout = TRUE, stderr = TRUE)}, 
                    error = function(e) {
                      paste("hisat2", "indexing failed with error:", e$message)
                      })
  
 tryCatch({system2('samtools', 
                           args = c('view', '-bS', meta_data$sam1[i], '-o', meta_data$bam1[i]),
                   stdout = TRUE, stderr = TRUE)}, 
                   error = function(e) {
                             paste("samtools", "sam>bam failed with error:", e$message)
                           })
  
  tryCatch({system2('samtools', 
                           args = c('sort', meta_data$bam1[i], '-o', meta_data$bam1sorted[i]),
                    stdout = TRUE, stderr = TRUE)}, 
                    error = function(e) {
                      paste("samtools", "sorting failed with error:", e$message)
                    })
  
  tryCatch({system2('samtools', 
                           args = c('index', meta_data$bam1sorted[i]),
                    stdout = TRUE, stderr = TRUE)}, 
                    error = function(e) {
                      paste("samtools", "indexing failed with error:", e$message)
                  })
  writeLines(hisat2,meta_data$hisatLog[i])
}
```

#question 5
```{r q5, warning = FALSE, message = FALSE}

# get the folder
hisat2_logs_dir <- "bam_files/"

# 1. Get a list of all HISAT2 log files
log_files <- list.files(hisat2_logs_dir, pattern = "*.log", full.names = TRUE)

#preper an "empty" vector of the apropreate size, faster than appending. 
percent_aligned <- 1:length(log_files)

#loop through log files
for (i in seq_along(percent_aligned)) {
  
  percent_aligned[i] <- readLines(log_files[i])[length(readLines(log_files[i]))]
  
}

#bind vectors as dataframe
align_df <- data.frame(sort(meta_data$SampleName),percent_aligned)

# extract the numeric percent value
align_df <- align_df %>% 
  mutate(percent_aligned = as.numeric(
    stringr::str_split_i(align_df$percent_aligned, "%",1))) %>% 
  rename(samplename = sort.meta_data.SampleName.)

head(align_df)

#question 5
#display align_df
DT::datatable(align_df)

#plot the data
ggplot(align_df, aes(x = "", y = percent_aligned)) +  
geom_boxplot() +
  xlab("") +  
  ylab("percent aligned") +
  ggtitle("Boxplot of alignment")

```

#Question 6
```{r q6, warning = FALSE, message = FALSE}
#list bam files
bfiles <- list.files("./bam_files", pattern = ".sorted.bam$", full.names = TRUE)

gene_count_list <- Rsubread::featureCounts(files = bfiles, annot.ext = "data/GCF_000001735.4_TAIR10.1_genomic.gtf", 
                                           isGTFAnnotationFile = TRUE, # <--- input annotation is GTF
                                           allowMultiOverlap = FALSE,  # <--- don't allow reads that overlap with multiple loci
                                           isPairedEnd = FALSE, nthreads = 8,
                                           minMQS = 10, # <--- minimum mapping quality score of 10 (like a phred score for the hisat2 alignment)
                                           GTF.featureType = "exon",  # <--- Count reads overlapping 'exon' features
                                           GTF.attrType = "gene_id" # <--- Groups exon by gene id
)
count_table <- gene_count_list$counts

#drop .sorted.bam from column titles
colnames(count_table) <- gsub(".sorted\\.bam$", "", colnames(count_table))

#filter out genes not mapped to any reads
filtered_table <- count_table[rowSums(count_table) != 0, ]

#interactable table
DT::datatable(filtered_table)
```

#Question 7
```{r q7, warning = FALSE, message = FALSE}
#Get the metadata table that would accompany the count table
coldata <- meta_data %>% dplyr::select(SampleName,SampleLong,Factor) %>% 
  dplyr::mutate(SampleLong=str_split_i(SampleLong, "\\.",1)) %>% #getting the groups name (Avirulent, Mock and Virulent)
  dplyr::rename(condition = SampleLong) %>%
  dplyr::mutate(condition = factor(condition)) %>% #for group comp
  dplyr::mutate(Factor = factor(Factor)) #for sample comp

base::rownames(coldata) <- coldata$SampleName
coldata <- coldata %>% mutate(SampleName = factor(SampleName))

coldata$type <- factor(rep("single-read", nrow(coldata)))

#Make sure samples are in the same order between the two. 
coldata <- coldata[base::match(base::colnames(count_table), rownames(coldata)),]

#Check that they are indeed same order
all(rownames(coldata) == base::colnames(count_table))

#creating a dds object where the conditions are avr vs mock vs vir

#creating a dds object where the conditions are the samples themslevs 
dds2 <- DESeqDataSetFromMatrix(countData = count_table,
                               colData = coldata,
                               design = ~ Factor)
dds2_results <- DESeq(dds2)
comp <- systemPipeR::readComp("/Users/mtaillefer00/Library/R/arm64/4.4/library/systemPipeRdata/extdata/param/targetsPE.txt")
comp[[1]]
compPair <- comp[[1]]

# Function to filter and count DE genes
filter_and_count <- function(res_obj, comparison_name, fc_threshold = 2) {
  # Remove NAs
  res_filtered <- res_obj[!is.na(res_obj$padj) & !is.na(res_obj$log2FoldChange), ]
  
  # Apply filters: |log2FC| >= log2(2) = 1 and padj <= alpha (already set in results())
  sig_genes <- res_filtered[abs(res_filtered$log2FoldChange) >= log2(fc_threshold), ]
  
  # Count up and down regulated
  up_regulated <- sum(sig_genes$log2FoldChange > 0)
  down_regulated <- sum(sig_genes$log2FoldChange < 0)
  
  return(data.frame(
    Comparison = comparison_name,
    Up_regulated = up_regulated,
    Down_regulated = down_regulated
  ))
}

results <- list()
for (pair in compPair){
  groups <- unlist(strsplit(pair,'-'))
  gr1 <- groups[1]
  gr2 <- groups[2]
  compName <- paste(gr1, 'vs', gr2)
  res <- DESeq2::results(dds2_results, contrast = c("Factor", gr1, gr2), alpha = 0.2)
  filtered_data <- filter_and_count(res, compName)
  results[[compName]] <- filtered_data
}

results_summary2 <- do.call(rbind, results)

plot_data2 <- results_summary2 %>%
  pivot_longer(cols = c(Up_regulated, Down_regulated),
               names_to = "Regulation",
               values_to = "Count") %>%
  mutate(Regulation = factor(Regulation, levels = c("Up_regulated", "Down_regulated")))

p_2 <- ggplot(plot_data2, aes(x = Comparison, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", position = "stack") +
  coord_flip() +
  labs(
    title = "Differentially Expressed Genes by Sample Comparison dd2",
    subtitle = "Fold Change >= 2,, alpha = 0.2",
    x = "Comparison",
    y = "Number of Genes",
    fill = "Regulation"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10))

p_2
```

#Question 8
When calculating Spearman correlation using all genes, we did not observe samples of the same type clustering together, whereas we do observe clustering of related samples when we calculated the correlation only using differentially expressed genes. This is since only a small proportion of genes are differentially expressed, so they had little effect on the correlation of all genes. When observing only differentially expressed genes, we can see how the plants are reacting to the different conditions. This matches what we expect biologically. 

